Reasoning Graphically about Quantum Computation

### 2008-03-05 papers/reasoning-graphically/paper.tex
more spider pictures
git-svn-id: https://isaplanner.svn.sourceforge.net/svnroot/isaplanner/trunk/quantomatic@1400 f8f375fb-4f10-0410-8d47-b6fbdc7ca053
work...
git-svn-id: https://isaplanner.svn.sourceforge.net/svnroot/isaplanner/trunk/quantomatic@1401 f8f375fb-4f10-0410-8d47-b6fbdc7ca053

\begin{figure}[t]
\includegraphics{images/gen-spider} = \includegraphics{images/gen-spider}
\label{fig:informal-spider-law}\caption{An informal equation on graphs
  that expresses the Spider Theorem. }
\end{figure}

@@

\noindent This motivates a characterisation that has only one
representative for such a set of equations. More presicely, we
formalise a representation that abstracts over the boundary nodes
membership in the domain or co-domain. This gives rise to a
\emph{variable-node graphs}, in which boundary nodes have been
generalised to \emph{variable nodes}. The intuition of variable nodes
is that they can replaced by concrete nodes in some graph. This is a
process analgous to composition. We now define the semantics for
variable-node graphs and then go on and introduce a way to join them.

% relationship to composition on the underlying concrete graphs.

\subsection{Semantics by Matching as an Open Embedding}

@@

\subsection{Variable-node Plugging}

While concrete graphs can be composed sequentially as well as tensored
in an adjacent manner, variable-node graphs enjoy a single but
somewhat wilder joining operation which we call \emph{plugging}.

\begin{definition}
  A \emph{plugging}, written $\pi_{p}(G,H)$, is a pair, $p$, of strict
  partial graph morphisms, $f_G : G \rightarrow H$ and $f_H : H
  \rightarrow G$ such that $f_G$ and $f_H$ \emph{agree}, following
  the definition given below, on node mappings and do not map concrete
  nodes.
\end{definition}

\begin{definition}
  Two mappings $f_G$ and $f_H$ \emph{agree} when $\forall v \in
  f_G(G).\, f_H(v) \in f^{-1}_G(v)$ and $\forall v \in f_H(H).\,
  f_G(v) \in f^{-1}_H(v)$. This states that if a set of nodes in one
  graph all map to a common node, $n$, in the other graph, then $n$
  maps back to one of the nodes that maps to it.
\end{definition}

We let $\pi(G,H)$ abbreviate $\exists p.\,\pi_p(G,H)$. We also
overload the notation $\pi_p(G,H)$ to denote the graph that results
from the plugging. This is defined to be the union of the nodes and
edges from $G$ and $H$ where the nodes and edges mapped by the pair of
plugging function are identified. Identification of nodes instantiates
variable nodes: when a variable node is mapped to a concrete one,
identification results in the concrete node.

While this is simply the intuitive way to combine graphs with variable
nodes, the definition takes some care. In particular, it is important
that plugging preserves incident edges of concrete nodes.
Figure~\ref{plugging-examples-fig} presents some examples to help the
reader get the intuition as well as understand some of the more
subtlte cases.

\begin{theorem}
  \emph{Plugging is symmetric}: $\forall p.\, \exists q.\, \pi_p(G,H)
  = \pi_q(H,G)$. Proof, given $p = (f_G,f_H)$ then $q = (f_H,f_G)$. As
  a corollary, we get $\pi(G,H) \Leftrightarrow \pi(H,G)$.
\end{theorem}

\begin{theorem}
  \emph{Plugging preserves matching}: $V \leq_v G \Rightarrow V \leq_v
  \pi_f(G,H)$. As corrolaries we get $V \leq_v H \Rightarrow V \leq_v
  \pi_f(G,H)$ from symmetry of plugging, $\vinterp{G} \cap \vinterp{H}
  \subseteq \bigcup_f \vinterp{\pi_f(G,H)}$ from the definition of
  interpretation, $G \leq_v \pi_f(G,H)$ from reflexivity of matching
  and consequently $H \leq_v \pi_f(G,H)$ by symmetry of plugging.
\end{theorem}

@@

A crucial feature of rules like the Spider Theorem is that an arbirary
number of repetitions of a subgraph need to be matched. To support
this we introduce an operation, !-boxing, on graph
representations. Given a graph representation, this introduces a new
notation by outlining (called !-boxing) a set of chosen
nodes. Intuitively, the resulting !-boxed graph can be thought of as
representing a set of graphs with an arbirary number of copies of the
!-boxed nodes, where every copy connects, in the same way, to the
nodes outside the !-box. The !-boxes can also be instantiated with
zero copies, which erases all edges to and from the !-box. See
Figure~\ref{spider-bang-rule} .

More formaly, a graph with !-boxes is a graph paired with a set
!-boxes. Each of these !-boxes is the set of nodes inside it and each
!-box must be disjoint from the others.\footnote{The disjointness
  condition avoids a slightly more complex, although also more
  expressive, semantics for overlaping node sets in the !-boxes. While
  such expressivity is interesting from a representational point of
  view, it is not needed to express the currently known rules of
  quantum computation.}

For the purposes of this paper, the underlying graph representation of
!-boxed graphs is variable-node graphs. The set of variable-node
graphs for a graph $G$ with !-boxes, is denoted by $\binterp{G}$. We
now define the concept of !-box matching which we use to provide a
semantics for the !-boxed graphs.

@@

To formalise the intuitive notion that a !-box represents arbitrary
number of copies of the subgraph made from it's nodes, we introduce
\emph{!-box matching}. This is a binary relation on graphs, written
infix as $G \leq_! H$ when $G$ matches $H$. The set of !-box graphs
matched by a !-box graph is defined as the set closed by the following
operations on the !-boxes in a graph:

@@

\begin{theorem}
  \emph{!-Matching respects !-box semantics}: $G \leq_! H
  \Leftrightarrow \binterp{G} \supseteq \binterp{H}$. The proof is a
  simple consequence from the defintion of $\binterp{G}$ being a
  subset of the graphs that match $G$.
\end{theorem}

@@

The representation of Compact Closed Categories as Graphs, discussed
in~\S\ref{graph-repr-comp}, is too restrictive for reasoning about
quantum computation. In particular, pictures such the Spider Theorem, in
Figure~\ref{spider-law-fig} are frequently needed, but not
expressible. Such pictures are used to denote an infinite family of
concrete graphs that can be rewritten. In this section we combine the
{\em !-boxes} with {\em variable nodes} graph extensions to form a
representation that allows us to finitely express these infinite
families of concrete graphs. We call this extension \ref{graph
  patterns}. The Spider Law can now be represented by the equation in
Figure~\ref{graph-pattern-spider-law-fig}. We also extend the notion
of matching for graph pattens as well as present a corresponding
notion of plugging. This provides the foundations for the rewriting
machinery in~\S\ref{rewriting} which can then be used to reasoning
about quantum computation.

graph-pattern-spider-law-fig

@@

\subsubsection{Completeness.}

Graph patterns are complete in the sense that for every concrete graph
there is a pattern that matches it. This is a trivial consequence of
the completeness of variable-node extension to graphs, which is
unaffected by allowing !-boxes.

@@

Because viable-node matching preserves the semantics, it is trivial to
prove that this definition meets the specification. This gives us
matching between pattern-nodes.

\subsubsection{Graph Pattern Equality.}

Equality of graph patterns can simply be defined as mutual matching:

\begin{definition}
  $G = H \Leftrightarrow G \leq H \land H \leq G$
\end{definition}

\subsection{Plugging}

We then extend the plugging
operation on variable-node graphs to graphs with !-boxes.

By defining graph patterns as a conservative extension of concrete
graphs we get the relative soundness for them.

@@

Axioms defined by an object-level model, as well as derived rules in
our framework, are pairs of pattern graph. The pair represents the
left and right hand sides of an equation. Rules are declarative in
that they denote set of concrete equational rules.

Substituon with a rule involves replacing a subgraph that matches the
left hand side with the right hand side. To do this there has to be a
isomorphism between variable nodes in the left and right hand
subgraphs. More concretely, there must be a unique naming of the
variable nodes in the left hand side such that each name occurs in a
unique variable node on the right hand side. Given a matching on the
left hand side, the target subgraph is replaced with the right hand
side while keeping the same instantiations for the isomorphic variable
nodes.

Rules also contain a partial 1-1 mapping between !-boxes on the left
and right hand sides. The intuition for this mapping is that the
unfolding used when matching, for a !-box on the left, is applied to
the mapped !-box on the right before replacement.

Notationally, and implementationally, we annotate !-boxes and variable
nodes in a graphs with unique names. For example,
Figure~\ref{spider-law2-fig} illustrates the Spider Law as a rule in
the general framework. The partial 1-1 mapping for !-boxes is
represented by !-boxes having the same name on the left and right of a
rule. Similarly, the isomorphism between variable nodes is captured by
the set of variable node names on the left and right hand side being
equal. There is an additional restriction that comes from the
interplay between !-boxes and variable nodes. When a variable node
appears within a !-box on one side of a rule it must appear under
mapped !-box on the other side. This enforces that rules cannot change
the type of a graph pattern.

\subsection{Adequacy}

This
provides rewriting for pattern graphs.

A an equation on a graph pattern is

\emph{Adequacy} says that if one by doing so within an
equation, we are not

@@

The meta-logical rules of the framework are quite simple as they only
involve dealing with object-level equations:

@@

\begin{center}
\prooftree
\Gamma \vdash A = B
\justifies
\Gamma \vdash B = A
\using\mbox{symmetry}
\endprooftree
\quad\quad
\prooftree
\Gamma \vdash A = B
\Gamma \vdash C = D
\justifies
\Gamma \vdash (C = D[A/B])
\using\mbox{substitution}
\endprooftree
\end{center}

@@

The proof of adequancy ...

The completeness of the resulting system, with respect to the intended
underlying semantics, depends on the completeness of the set of given
rules.

\subsection{Derived Rules}

\subsection{Simplification}

\subsection{Adequacy}

@@

\section{A Case Study in Quantum Computation}

The model of qantum computation in which we are interested is finite
dimensional hilbert spaces.

@@

Completeness is a challenging and still open problem for graph based
models of quantum computation. However, even without completeness
interesting quantum computations can be expressed.

>>>>

\begin{figure}[t]
$$\begin{array}{ccc}
\inlinegraphic{6em}{spider_lhs} & = & \inlinegraphic{6em}{spider_rhs}
\end{array}$$
\label{fig:informal-spider-law}\caption{An informal equation on graphs
  that expresses the Spider Theorem. }
\end{figure}

@@

\noindent This motivates a characterisation that has only one
representative for such a set of equations. More presicely, we
formalise a representation that abstracts over the boundary nodes
membership in the domain or co-domain. This gives rise to a
\emph{variable-node graphs}, in which boundary nodes have been
generalised to \emph{variable nodes}. The intuition of variable nodes
is that they can replaced by concrete nodes in some graph in a process
analgous to composition.

% We now define the semantics for variable-node
% graphs and then go on and introduce a way to join them.
% relationship to composition on the underlying concrete graphs.

% \subsection{Semantics by Matching as an Open Embedding}

@@

% \subsection{Variable-node Plugging}

% While concrete graphs can be composed sequentially as well as tensored
% in an adjacent manner, variable-node graphs enjoy a single but
% somewhat wilder joining operation which we call \emph{plugging}.

% \begin{definition}
%   A \emph{plugging}, written $\pi_{p}(G,H)$, is a pair, $p$, of strict
%   partial graph morphisms, $f_G : G \rightarrow H$ and $f_H : H
%   \rightarrow G$ such that $f_G$ and $f_H$ \emph{agree}, following
%   the definition given below, on node mappings and do not map concrete
%   nodes.
% \end{definition}

% \begin{definition}
%   Two mappings $f_G$ and $f_H$ \emph{agree} when $\forall v \in
%   f_G(G).\, f_H(v) \in f^{-1}_G(v)$ and $\forall v \in f_H(H).\,
%   f_G(v) \in f^{-1}_H(v)$. This states that if a set of nodes in one
%   graph all map to a common node, $n$, in the other graph, then $n$
%   maps back to one of the nodes that maps to it.
% \end{definition}

% We let $\pi(G,H)$ abbreviate $\exists p.\,\pi_p(G,H)$. We also
% overload the notation $\pi_p(G,H)$ to denote the graph that results
% from the plugging. This is defined to be the union of the nodes and
% edges from $G$ and $H$ where the nodes and edges mapped by the pair of
% plugging function are identified. Identification of nodes instantiates
% variable nodes: when a variable node is mapped to a concrete one,
% identification results in the concrete node.

% While this is simply the intuitive way to combine graphs with variable
% nodes, the definition takes some care. In particular, it is important
% that plugging preserves incident edges of concrete nodes.
% Figure~\ref{plugging-examples-fig} presents some examples to help the
% reader get the intuition as well as understand some of the more
% subtlte cases.

% \begin{theorem}
%   \emph{Plugging is symmetric}: $\forall p.\, \exists q.\, \pi_p(G,H)
%   = \pi_q(H,G)$. Proof, given $p = (f_G,f_H)$ then $q = (f_H,f_G)$. As
%   a corollary, we get $\pi(G,H) \Leftrightarrow \pi(H,G)$.
% \end{theorem}

% \begin{theorem}
%   \emph{Plugging preserves matching}: $V \leq_v G \Rightarrow V \leq_v
%   \pi_f(G,H)$. As corrolaries we get $V \leq_v H \Rightarrow V \leq_v
%   \pi_f(G,H)$ from symmetry of plugging, $\vinterp{G} \cap \vinterp{H}
%   \subseteq \bigcup_f \vinterp{\pi_f(G,H)}$ from the definition of
%   interpretation, $G \leq_v \pi_f(G,H)$ from reflexivity of matching
%   and consequently $H \leq_v \pi_f(G,H)$ by symmetry of plugging.
% \end{theorem}

@@

A crucial feature of rules like the Spider Theorem is that an arbirary
number of repetitions of a subgraph need to be matched. To support
this we introduce an operation, !-boxing (pronounced band-boxing), on
graph representations. Given a graph representation, this introduces a
new notation by outlining a set of nodes. These nodes are said to be
in a !-box. Intuitively, the resulting !-boxed graph can be thought of
as representing a set of graphs with an arbirary number of copies of
the !-boxed nodes, where every copy connects, in the same way, to the
nodes outside the !-box. The !-boxes can also be instantiated with
zero copies. This erases all edges to and from the !-box.

%Figure~\ref{fig:spider-bang-lhs-example} shows an how the left hand side of
%the spider theorem can be represented with bang boxes and illustrates
%some instances of it.

%\begin{figure}[t]
%$$\begin{array}{ccc}
%\inlinegraphic{6em}{spider_lhs_patt} & = & \inlinegraphic{6em}{spider_rhs_ex1} & , & \inlinegraphic{6em}{spider_rhs_ex2} & , & \inlinegraphic{6em}{spider_rhs_ex3} & \dots
%\end{array}$$
%\label{fig:spider-bang-lhs-example}\caption{An informal equation on graphs
%  that expresses the Spider Theorem. }
%\end{figure}

More formaly, a !-box graph is a graph paired with a set !-boxes. Each
of these !-boxes is the set of nodes inside it and each !-box must be
disjoint from the others.\footnote{The disjointness condition avoids a
  slightly more complex, although also more expressive, semantics for
  overlaping node sets in the !-boxes. While such expressivity is
  interesting from a representational point of view, it is not needed
  to express the currently known rules for quantum computation.}  We
now define the concept of !-box matching which we then use to provide
a formal semantics for the !-boxed graphs.

@@

To formalise the intuitive notion that a !-box represents arbitrary
number of copies of the subgraph made from it's nodes, we introduce
\emph{!-box matching}. This is a binary relation on graphs, written
infix as $G \leq_! H$ when $G$ matches $H$. The set matching !-box
graphs is defined as the set closed by the following operations on the
!-boxes in a graph:

@@

\begin{theorem}
\label{thm:bang-box-respect}
\emph{!-Matching respects !-box semantics}: $G \leq_! H
\Leftrightarrow \binterp{G} \supseteq \binterp{H}$. The proof is a
simple consequence from the defintion of $\binterp{G}$ being a subset
of the graphs that match $G$.
\end{theorem}

For the purposes of this paper, the underlying graph representation of
!-boxed graphs is variable-node graphs. This gives rise to graph
patterns which we now discuss in more detail.
% The set of variable-node
% graphs for a graph $G$ with !-boxes, is denoted by $\binterp{G}$.

@@

The representation of Compact Closed Categories as Graphs, discussed
in~\S\ref{sec:graph-repr-comp}, is too restrictive for reasoning about
quantum computation. In particular, graphical rules such the Spider
Theorem, from Figure~\ref{fig:informal-spider-law}, are frequently
needed, but not expressible. In this section we combine the {\em
  !-boxes} and {\em variable-node} extensions to graphs. We call this
representation \emph{graph patterns}. This forms a representation that
allows us to express, in a finite way, the infinite family of concrete
graphs equations. In particular, the Spider Theroem can now be
represented by the equation shown in Figure~\ref{fig:spider-thm-patt}.
We also extend the notion of matching for graph pattens.
%as well as present a corresponding notion of plugging. 
This provides the foundations for the rewriting machinery
in~\S\ref{sec:rewriting} which can then be used to reasoning about
quantum computation.

\begin{figure}[t]
$$\begin{array}{ccc}
\inlinegraphic{8em}{spider_lhs_patt} & \;=\; & \inlinegraphic{6em}{spider_rhs_patt}
\end{array}$$
\label{fig:spider-thm-patt}\caption{The Spider Theorem expressed using graph patterns.}
\end{figure}

% \subsubsection{Completeness.}

% Graph patterns are complete in the sense that for every concrete graph
% there is a pattern that matches it. This is a trivial consequence of
% the completeness of variable-node extension to graphs, which is
% unaffected by allowing !-boxes.

@@

@@

Because variable-node matching preserves the semantics, it is trivial to
prove that this definition meets the specification.

% \subsubsection{Graph Pattern Equality.}

% Equality of graph patterns can simply be defined as mutual matching:

% \begin{definition}
%   $G = H \Leftrightarrow G \leq H \land H \leq G$
% \end{definition}

% \subsection{Plugging}

% We then extend the plugging
% operation on variable-node graphs to graphs with !-boxes.

% By defining graph patterns as a conservative extension of concrete
% graphs we get the relative soundness for them.

% We note that using concrete graphs as the underlying representation
% results in composition of !-box graphs to disrespect the !-box
% semantics: Theorem \ref{thm:bang-box-respects-plugging} fails to hold
% for composision.

@@

Axioms defined by an object-level model, as well as derived rules in
our framework, are pairs of pattern graph. The pair represents the
left and right hand sides of an equation. Rules are declarative in
that they denote a set of concrete equational rules.

The intuitive idea of substituon with a rule is to replace a subgraph
that matches the left hand side with the right hand side. However, not
all pairs of rules are make sense with respect to the underlying
semantics. For an equation to be well defined with respect to the
compact closed structure it must not be possible to changes the type
of a concrete graph by rewriting. Mapping this restriction back to
pairs of graph patterns results in the following conditions on rules:

\begin{itemize}

\item There has to be a isomorphism between variable nodes in the left
  and right hand subgraphs. More concretely, there must be a unique
  naming of the variable nodes in the left hand side such that each
  name occurs in a unique variable node on the right hand side. Given
  a matching on the left hand side, the target subgraph is replaced
  with the right hand side while keeping the same instantiations for
  the isomorphic variable nodes.

\item Rules must also define a partial 1-1 mapping between !-boxes on
  the left and right hand sides. The intuition for this mapping is
  that the unfolding used when matching a !-box on the left, is
  applied to the mapped !-box on the right before replacement.

\item The interplay between !-boxes and variable nodes means that when
  a variable node appears within a !-box on one side of a rule it must
  also appear under a mapped !-box on the other side.

Notationally, and implementationally, we annotate !-boxes and variable
nodes in a graphs with unique names. For example, see
Figure~\ref{fig:spider-thm-patt} which illustrates the Spider Theorem
as a rule between graph patterns. In this figure, the partial 1-1
mapping for !-boxes is represented by !-boxes having the same name on
the left and right of a rule.  Similarly, the isomorphism between
variable nodes is captured by the set of variable node names on the
left and right hand side being equal.

\subsubsection{Adequacy.} An equation on graph patterns corresponds to
a possibly infinite set of equations betweens concrete graphs. We
might expect that such equations express too much: they may allow
rewrites which are not intended. We prove \emph{adequacy}, which
states that this is not the case. In particular, given an equation
between concrete graphs, $G = H$, the equations between the lifting of
these to variable node graphs by replacing boundary nodes with
variable nodes is a conservative extension. The proof follows directly
from the obervation that every instance of the lifted variable node
equation corresponds to compsition of $G$ and $H$ with some other
graph (the non-embedded part).

@@

Having defined what make a valid rule, we now present the meta-logic
of the framework. This is quite simple as they only involve dealing
with object-level equations:

@@

\begin{center}
\prooftree
\Gamma \vdash A = B
\justifies
\Gamma \vdash B = A
\using\mbox{symmetry}
\endprooftree
\quad\quad
\prooftree
\Gamma \vdash A = B \quad\quad
\Gamma \vdash C = D
\justifies
\Gamma \vdash (C = D[A/B])
\using\mbox{substitution}
\endprooftree
\end{center}

@@

The completeness of the resulting system, with respect to the intended
underlying semantics, depends on the completeness of the set of given
rules. Completeness is a challenging and still open problem for graph
based models of quantum computation. Even without completeness, we
obersve in \S\ref{sec:case-study} that interesting and useful quantum
computations can be expressed.

\section{Implementation}
\label{ssec:implementation}

% \subsection{Simplification}

% One of the main applications of

@@

\section{A Case Study in Quantum Computation}
\label{sec:case-study}
%The model of qantum computation in which we are interested is finite
%dimensional hilbert spaces.

@@

\subsection{}
\label{ssec:node-expressions}
